---
title: "RAP environment check"
description: "Check whether UKBAnalytica is running in an approved UK Biobank RAP environment."
---

UKBAnalytica is designed for analyses conducted within approved UK Biobank Research Analysis Platform (RAP) projects. Participant-level UK Biobank source records should remain inside RAP. The package therefore provides a lightweight environment check and protects RAP-dependent extraction interfaces from being run in ordinary local sessions.

::: {.callout-important title="Privacy boundary"}
UKBAnalytica can help users write scripts, define phenotypes, assemble analysis workflows, and summarize derived results. It should not be used to move or expose UK Biobank participant-level source records outside approved RAP projects.
:::

## Check the current session

Use `ukb_check_rap_env()` at the beginning of RAP scripts to inspect the current execution environment.

```{r}
library(UKBAnalytica)

env <- ukb_check_rap_env()
```

The function returns an object with structured diagnostics:

```{r}
env <- ukb_check_rap_env(verbose = FALSE)

env$is_rap
env$rap_signals
env$project_mount_exists
env$dx_available
env$dx_path
env$dx_version
env$auth
env$checks
```

The main RAP signals are:

- `DX_PROJECT_CONTEXT_ID`, `DX_WORKSPACE_ID`, `DX_JOB_ID`, or `DX_RUN_ID`
- the RAP project mount path `/mnt/project`
- the DNAnexus command-line tool `dx`, when available

If at least one RAP environment signal is detected, `env$is_rap` is returned as `TRUE`. The `checks` table records each diagnostic item as `pass`, `warn`, `fail`, or `skip`.

## Require RAP explicitly

For scripts that should only run inside RAP, set strict checks:

```{r}
ukb_check_rap_env(
  require_rap = TRUE,
  require_dx = TRUE
)
```

To also verify DNAnexus authentication:

```{r}
ukb_check_rap_env(
  require_rap = TRUE,
  require_dx = TRUE,
  check_auth = TRUE
)
```

`check_auth = TRUE` calls `dx whoami` and `dx env --bash` to verify the DNAnexus login and active project context. It does not read participant-level data.

## Protected RAP functions

Some UKBAnalytica functions are intentionally restricted because they interact with RAP datasets or RAP project storage. For example:

```{r}
rap_extract_pheno(field_id = c(31, 53, 21022))
rap_submit_extract(field_id = c(31, 53, 21022), file = "baseline_core")
```

These functions now call an internal RAP assertion before running the actual extraction step. If they are executed outside a RAP-like environment, they stop with a diagnostic message and ask the user to run `ukb_check_rap_env()`.

Dry-run planning remains available without extracting data:

```{r}
plan <- rap_submit_extract(
  field_id = c(31, 53, 21022),
  dataset = "appXXXX_YYYYMMDD.dataset",
  dry_run = TRUE
)
```

This allows users to review field selections and extraction plans without touching participant-level records.

## Recommended script header

A typical RAP analysis script can begin with:

```{r}
library(UKBAnalytica)

rap_env <- ukb_check_rap_env(
  require_rap = TRUE,
  require_dx = TRUE,
  check_auth = TRUE
)
```

This makes the privacy boundary explicit and helps prevent accidental local execution of RAP-dependent workflows.
