# Main analysis {#main-analysis}

This chapter shows the core statistical-analysis functions in
**UKBAnalytica**. The main workflow is intentionally simple: first inspect
correlations among candidate variables, then run one of three regression
families according to the outcome type.

## Correlation pre-analysis {#correlation}

Before regression modeling, `run_correlation()` can be used to inspect
correlations among candidate exposures or covariates. This is useful for
identifying highly correlated variables before building adjusted models.

```{r correlation-basic, eval=FALSE}
library(UKBAnalytica)

corr_mat <- run_correlation(
  df = analysis_data,
  vars = c("age", "bmi", "sbp", "dbp", "glucose", "hba1c"),
  method = "pearson",
  threshold = 0.70
)

corr_mat
```

Use `plot_correlation()` when a visual summary is needed.

```{r correlation-plot, eval=FALSE}
plot_correlation(
  corr_mat,
  show_values = TRUE,
  upper_triangle = TRUE,
  title = "Correlation matrix"
)
```

Common choices are `method = "pearson"` for approximately continuous variables
and `method = "spearman"` for rank-based or skewed variables.

## Regression analyses {#regression}

The three main regression functions share the same design: each variable in
`main_var` is tested separately, and `covariates` defines the adjustment set.
This makes it straightforward to screen multiple exposures while keeping model
specification consistent.

| Outcome type | Function | Effect estimate |
|---|---|---|
| Continuous outcome | `runmulti_lm()` | beta |
| Binary outcome | `runmulti_logit()` | odds ratio (OR) |
| Time-to-event outcome | `runmulti_cox()` | hazard ratio (HR) |

For high-dimensional analyses, such as biomarker or protein screening, adjust
the returned p-values with `p.adjust()`.

```{r p-adjust, eval=FALSE}
results$p_bh <- p.adjust(results$pvalue, method = "BH")
results$p_bonferroni <- p.adjust(results$pvalue, method = "bonferroni")
```

### Linear regression

Use `runmulti_lm()` for continuous outcomes.

```{r lm-example, eval=FALSE}
lm_res <- runmulti_lm(
  data = analysis_data,
  main_var = c("bmi", "smoking_status", "hba1c"),
  outcome = "sbp",
  covariates = c("age", "sex", "ethnic")
)

lm_res
```

The output includes `variable`, `beta`, `lower95`, `upper95`, and `pvalue`.
Here, `beta` represents the adjusted mean difference in the continuous outcome
per unit increase or category contrast of the tested variable.

### Logistic regression

Use `runmulti_logit()` for binary outcomes coded as 0/1.

```{r logit-example, eval=FALSE}
logit_res <- runmulti_logit(
  data = analysis_data,
  main_var = c("bmi", "smoking_status", "hba1c"),
  outcome = "diabetes_status",
  covariates = c("age", "sex", "ethnic")
)

logit_res
```

The output includes `variable`, `OR`, `lower95`, `upper95`, and `pvalue`.
The outcome must be binary, with 1 indicating the event or case group.

### Cox proportional hazards regression

Use `runmulti_cox()` for incident outcomes with follow-up time.

```{r cox-example, eval=FALSE}
cox_res <- runmulti_cox(
  data = survival_data,
  main_var = c("bmi", "smoking_status", "hba1c"),
  endpoint = c("follow_up_time", "incident_status"),
  covariates = c("age", "sex", "ethnic")
)

cox_res
```

The output includes `variable`, `coef`, `HR`, `lower95`, `upper95`, `pvalue`,
`n`, and `n_event`. The `endpoint` argument should contain the follow-up time
column followed by the event indicator column, where 1 indicates an incident
event and 0 indicates censoring.

## Minimal workflow

```{r minimal-workflow, eval=FALSE}
exposures <- c("bmi", "smoking_status", "hba1c")
adjustment <- c("age", "sex", "ethnic")

corr_mat <- run_correlation(
  analysis_data,
  vars = c(exposures, adjustment),
  method = "spearman",
  threshold = 0.70
)

cox_res <- runmulti_cox(
  data = survival_data,
  main_var = exposures,
  endpoint = c("follow_up_time", "incident_status"),
  covariates = adjustment
)

cox_res$p_bh <- p.adjust(cox_res$pvalue, method = "BH")
cox_res
```

This structure covers the main statistical layer of UKBAnalytica: correlation
pre-analysis followed by linear, logistic, or Cox regression according to the
outcome definition.
