# Main analysis {#main-analysis}

This chapter shows the core statistical-analysis functions in
**UKBAnalytica**. The main workflow is intentionally simple: first inspect
correlations among candidate variables, then run one of six regression
families according to the outcome type — either via the individual functions
or through the unified `run_regression()` interface.

## Correlation pre-analysis {#correlation}

Before regression modeling, `run_correlation()` can be used to inspect
correlations among candidate exposures or covariates. This is useful for
identifying highly correlated variables before building adjusted models.

```{r correlation-basic, eval=FALSE}
library(UKBAnalytica)

corr_mat <- run_correlation(
  df = analysis_data,
  vars = c("age", "bmi", "sbp", "dbp", "glucose", "hba1c"),
  method = "pearson",
  threshold = 0.70
)

corr_mat
```

Use `plot_correlation()` when a visual summary is needed.

```{r correlation-plot, eval=FALSE}
plot_correlation(
  corr_mat,
  show_values = TRUE,
  upper_triangle = TRUE,
  title = "Correlation matrix"
)
```

Common choices are `method = "pearson"` for approximately continuous variables
and `method = "spearman"` for rank-based or skewed variables.

## Regression analyses {#regression}

The six main regression functions share the same design: each variable in
`main_var` is tested separately, and `covariates` defines the adjustment set.
This makes it straightforward to screen multiple exposures while keeping model
specification consistent.

| Outcome type | Function | Effect estimate |
|---|---|---|
| Continuous outcome | `runmulti_lm()` | beta (mean difference) |
| Binary outcome | `runmulti_logit()` | odds ratio (OR) |
| Time-to-event outcome | `runmulti_cox()` | hazard ratio (HR) |
| Count / rate outcome | `runmulti_glm()` | beta on link scale |
| Overdispersed count | `runmulti_negbin()` | incidence rate ratio (IRR) |
| Non-linear association | `runmulti_gam()` | edf + p-value (smooth) / beta (linear) |

For high-dimensional analyses, such as biomarker or protein screening, adjust
the returned p-values with `p.adjust()`.

```{r p-adjust, eval=FALSE}
results$p_bh <- p.adjust(results$pvalue, method = "BH")
results$p_bonferroni <- p.adjust(results$pvalue, method = "bonferroni")
```

### Linear regression

Use `runmulti_lm()` for continuous outcomes.

```{r lm-example, eval=FALSE}
lm_res <- runmulti_lm(
  data = analysis_data,
  main_var = c("bmi", "smoking_status", "hba1c"),
  outcome = "sbp",
  covariates = c("age", "sex", "ethnic")
)

lm_res
```

The output includes `variable`, `beta`, `lower95`, `upper95`, and `pvalue`.
Here, `beta` represents the adjusted mean difference in the continuous outcome
per unit increase or category contrast of the tested variable.

### Logistic regression

Use `runmulti_logit()` for binary outcomes coded as 0/1.

```{r logit-example, eval=FALSE}
logit_res <- runmulti_logit(
  data = analysis_data,
  main_var = c("bmi", "smoking_status", "hba1c"),
  outcome = "diabetes_status",
  covariates = c("age", "sex", "ethnic")
)

logit_res
```

The output includes `variable`, `OR`, `lower95`, `upper95`, and `pvalue`.
The outcome must be binary, with 1 indicating the event or case group.

### Cox proportional hazards regression

Use `runmulti_cox()` for incident outcomes with follow-up time.

```{r cox-example, eval=FALSE}
cox_res <- runmulti_cox(
  data = survival_data,
  main_var = c("bmi", "smoking_status", "hba1c"),
  endpoint = c("follow_up_time", "incident_status"),
  covariates = c("age", "sex", "ethnic")
)

cox_res
```

The output includes `variable`, `coef`, `HR`, `lower95`, `upper95`, `pvalue`,
`n`, and `n_event`. The `endpoint` argument should contain the follow-up time
column followed by the event indicator column, where 1 indicates an incident
event and 0 indicates censoring.

### Generalised linear models

Use `runmulti_glm()` when the outcome distribution does not fit linear or
logistic regression — for example, count outcomes (Poisson), overdispersed
counts (quasi-Poisson), or positive-continuous outcomes (Gamma).

The `family` argument accepts a character string, a family function, or a
family object, which allows the link function to be customised.

```{r glm-poisson, eval=FALSE}
# Poisson regression — incidence rate ratio = exp(beta)
poisson_res <- runmulti_glm(
  data = analysis_data,
  main_var = c("bmi", "smoking_status", "hba1c"),
  family = "poisson",
  outcome = "hospitalisation_count",
  covariates = c("age", "sex", "ethnic")
)

poisson_res
# exp(poisson_res$beta)      # incidence rate ratio
# exp(poisson_res$lower95)   # lower 95 % CI on ratio scale
# exp(poisson_res$upper95)   # upper 95 % CI on ratio scale
```

The output includes `variable`, `family`, `link`, `beta`, `lower95`,
`upper95`, `pvalue`, and `n`. Coefficients are on the link scale; for
log-link families exponentiate to obtain the ratio-scale effect.

For overdispersed count outcomes, switch to quasi-Poisson. Confidence
intervals are computed via the Wald method because profile-likelihood CIs are
not available for quasi-likelihood models.

```{r glm-quasipoisson, eval=FALSE}
qpois_res <- runmulti_glm(
  data = analysis_data,
  main_var = c("bmi", "smoking_status"),
  family = "quasipoisson",          # Wald CI used automatically
  outcome = "hospitalisation_count",
  covariates = c("age", "sex", "ethnic")
)
```

For positive-continuous outcomes such as biomarker concentrations, a Gamma
model with a log link is often preferable to a log-transformed linear model.

```{r glm-gamma, eval=FALSE}
gamma_res <- runmulti_glm(
  data = analysis_data,
  main_var = c("bmi", "smoking_status"),
  family = stats::Gamma(link = "log"),  # family object with custom link
  outcome = "crp",
  covariates = c("age", "sex", "ethnic")
)
```

### Negative-binomial regression

Use `runmulti_negbin()` for overdispersed count outcomes as a more principled
alternative to quasi-Poisson.  The model estimates the overdispersion
parameter `theta` alongside the effect; larger `theta` indicates less
overdispersion (approaching Poisson behaviour as `theta → ∞`).

```{r negbin-example, eval=FALSE}
negbin_res <- runmulti_negbin(
  data = analysis_data,
  main_var = c("bmi", "smoking_status", "hba1c"),
  outcome = "hospitalisation_count",
  covariates = c("age", "sex", "ethnic")
)

negbin_res
```

The output includes `variable`, `IRR` (incidence rate ratio = `exp(beta)`),
`lower95`, `upper95`, `pvalue`, `theta`, and `n`.  Confidence intervals use
profile likelihood.

**When to prefer negative-binomial over quasi-Poisson:** negative-binomial is
a proper probability model with a likelihood, enabling AIC-based model
comparison and more reliable inference under heavy overdispersion.
Quasi-Poisson adjusts the standard errors but does not change the coefficient
estimates.

### Generalised additive models (GAM)

Use `runmulti_gam()` to test whether exposures have non-linear associations
with the outcome.  By default each main variable enters the model as a
penalised thin-plate regression spline `s(var)`, allowing the dose-response
curve to take any smooth shape.

```{r gam-smooth, eval=FALSE}
gam_res <- runmulti_gam(
  data = analysis_data,
  main_var = c("bmi", "hba1c"),
  outcome = "sbp",
  covariates = c("age", "sex", "ethnic")
)

gam_res
```

When `smooth = TRUE` (default) the output includes `variable`, `edf`
(estimated degrees of freedom), `ref_df`, `stat` (F- or chi-squared
statistic depending on the response family), `stat_type`, `pvalue`, `family`,
`link`, and `n`.  An `edf` close to 1 suggests a near-linear relationship;
`edf > 1` indicates non-linearity.

A common workflow is to screen with the smooth GAM first, then re-fit a
linear model for variables where `edf ≈ 1`.

```{r gam-workflow, eval=FALSE}
# Step 1 — screen for non-linearity
gam_screen <- runmulti_gam(
  data = analysis_data,
  main_var = c("bmi", "hba1c", "crp", "age"),
  outcome = "sbp",
  covariates = c("sex", "ethnic")
)

# Variables with edf close to 1 are approximately linear
linear_vars  <- gam_screen$variable[gam_screen$edf < 1.5]
nonlin_vars  <- gam_screen$variable[gam_screen$edf >= 1.5]

# Step 2 — fit linear model for linear variables
lm_res <- runmulti_lm(
  data = analysis_data,
  main_var = linear_vars,
  outcome = "sbp",
  covariates = c("sex", "ethnic")
)

# Step 3 — keep GAM smooth for non-linear variables (visualise separately)
```

For count outcomes, specify the appropriate family.

```{r gam-poisson, eval=FALSE}
gam_count <- runmulti_gam(
  data = analysis_data,
  main_var = c("bmi", "hba1c"),
  outcome = "hospitalisation_count",
  family = "poisson",
  covariates = c("age", "sex", "ethnic")
)
# stat_type will be "Chi.sq" for Poisson (vs "F" for Gaussian)
```

To fit a parametric linear term inside a GAM (e.g. when non-linearity is
ruled out but GAM covariates are still desired), set `smooth = FALSE`.

```{r gam-linear, eval=FALSE}
gam_linear <- runmulti_gam(
  data = analysis_data,
  main_var = c("bmi", "hba1c"),
  outcome = "sbp",
  smooth = FALSE,
  covariates = c("age", "sex", "ethnic")
)
# Output mirrors runmulti_glm: beta, lower95, upper95, pvalue
```

## Unified interface: `run_regression()` {#run-regression}

`run_regression()` wraps all six regression functions behind a single call.
The `type` argument selects the model family; all other arguments are
forwarded to the corresponding underlying function.

| `type` | Underlying function | Key extra args |
|---|---|---|
| `"lm"` | `runmulti_lm()` | `outcome` |
| `"logit"` | `runmulti_logit()` | `outcome` |
| `"cox"` | `runmulti_cox()` | `endpoint` |
| `"glm"` | `runmulti_glm()` | `outcome`, `family` (default `"poisson"`) |
| `"negbin"` | `runmulti_negbin()` | `outcome` |
| `"gam"` | `runmulti_gam()` | `outcome`, `family`, `smooth` |

### Continuous outcome

```{r rr-lm, eval=FALSE}
run_regression(
  data       = analysis_data,
  main_var   = c("bmi", "smoking_status", "hba1c"),
  type       = "lm",
  outcome    = "sbp",
  covariates = c("age", "sex", "ethnic")
)
```

### Binary outcome

```{r rr-logit, eval=FALSE}
run_regression(
  data       = analysis_data,
  main_var   = c("bmi", "smoking_status", "hba1c"),
  type       = "logit",
  outcome    = "diabetes_status",
  covariates = c("age", "sex", "ethnic")
)
```

### Time-to-event outcome

```{r rr-cox, eval=FALSE}
run_regression(
  data       = survival_data,
  main_var   = c("bmi", "smoking_status", "hba1c"),
  type       = "cox",
  endpoint   = c("follow_up_time", "incident_status"),
  covariates = c("age", "sex", "ethnic")
)
```

### Count outcome (GLM)

```{r rr-glm, eval=FALSE}
run_regression(
  data       = analysis_data,
  main_var   = c("bmi", "smoking_status", "hba1c"),
  type       = "glm",
  family     = "poisson",            # default; change to "quasipoisson" etc.
  outcome    = "hospitalisation_count",
  covariates = c("age", "sex", "ethnic")
)
```

### Count outcome (negative-binomial)

```{r rr-negbin, eval=FALSE}
run_regression(
  data       = analysis_data,
  main_var   = c("bmi", "smoking_status", "hba1c"),
  type       = "negbin",
  outcome    = "hospitalisation_count",
  covariates = c("age", "sex", "ethnic")
)
```

### Non-linear screening (GAM)

```{r rr-gam, eval=FALSE}
# Smooth GAM (default) — tests association and non-linearity
run_regression(
  data       = analysis_data,
  main_var   = c("bmi", "smoking_status", "hba1c"),
  type       = "gam",
  outcome    = "sbp",
  covariates = c("age", "sex", "ethnic")
)

# Linear GAM — parametric term inside a GAM framework
run_regression(
  data       = analysis_data,
  main_var   = c("bmi", "hba1c"),
  type       = "gam",
  smooth     = FALSE,
  outcome    = "sbp",
  covariates = c("age", "sex", "ethnic")
)
```

`run_regression()` is convenient when the model type is determined
programmatically or when a single, consistent call site is preferred across
different analysis scripts.

## Minimal workflow

```{r minimal-workflow, eval=FALSE}
exposures  <- c("bmi", "smoking_status", "hba1c")
adjustment <- c("age", "sex", "ethnic")

# Pre-analysis: correlation screening
corr_mat <- run_correlation(
  analysis_data,
  vars      = c(exposures, adjustment),
  method    = "spearman",
  threshold = 0.70
)

# Non-linearity check with GAM before committing to linear Cox
gam_screen <- run_regression(
  data       = analysis_data,
  main_var   = exposures,
  type       = "gam",
  outcome    = "sbp",
  covariates = adjustment
)

# Primary analysis: Cox regression for incident events
cox_res <- run_regression(
  data       = survival_data,
  main_var   = exposures,
  type       = "cox",
  endpoint   = c("follow_up_time", "incident_status"),
  covariates = adjustment
)
cox_res$p_bh <- p.adjust(cox_res$pvalue, method = "BH")

# Secondary: negative-binomial for hospital admissions count
negbin_res <- run_regression(
  data       = analysis_data,
  main_var   = exposures,
  type       = "negbin",
  outcome    = "hospitalisation_count",
  covariates = adjustment
)
negbin_res$p_bh <- p.adjust(negbin_res$pvalue, method = "BH")
```

This structure covers the main statistical layer of UKBAnalytica: correlation
pre-analysis, non-linearity screening with GAM, followed by linear, logistic,
Cox, generalised linear, negative-binomial, or GAM regression according to the
outcome definition.
