# Advanced analysis {#advanced-analysis}

This chapter summarizes the advanced statistical modules in **UKBAnalytica**.
These functions are intended for common epidemiological extensions after the
main regression analyses: subgroup analysis, propensity score analysis, and
causal mediation analysis.

## Module overview

| Analysis task | Main functions | Typical use |
|---|---|---|
| Subgroup analysis | `run_subgroup_analysis()`, `run_multi_subgroup()` | Estimate effects within strata and report interaction p-values |
| Propensity score analysis | `estimate_propensity_score()`, `match_propensity()`, `calculate_weights()`, `run_weighted_analysis()` | Reduce confounding in observational treatment or exposure comparisons |
| Mediation analysis | `run_mediation()`, `run_multi_mediator()` | Decompose total effects into direct and indirect pathways |

For multiple imputation workflows, see the [Multiple imputation](06-imputation.qmd)
chapter.

## Subgroup analysis {#subgroup}

`run_subgroup_analysis()` estimates an interaction p-value from a full model
containing an exposure-by-subgroup term, then refits separate models within
each subgroup level and returns a harmonised table of effect estimates,
confidence intervals, and interaction statistics.

Five model families are supported via the `model_type` argument:

| `model_type` | Fitting function | Effect estimate |
|---|---|---|
| `"cox"` | `survival::coxph()` | Hazard ratio (HR) |
| `"logistic"` | `stats::glm(binomial)` | Odds ratio (OR) |
| `"linear"` | `stats::lm()` | Beta (mean difference) |
| `"glm"` | `stats::glm(family)` | Ratio (log/logit link) or beta (identity) |
| `"negbin"` | `MASS::glm.nb()` | Incidence rate ratio (IRR) |

### Time-to-event outcome

```{r subgroup-cox, eval=FALSE}
subgroup_res <- run_subgroup_analysis(
  data       = survival_data,
  exposure   = "smoking_status",
  subgroup_var = "sex",
  model_type = "cox",
  endpoint   = c("follow_up_time", "incident_status"),
  covariates = c("age", "bmi", "ethnic")
)

subgroup_res
```

### Count outcome (Poisson GLM)

Use `model_type = "glm"` with `family = "poisson"` for count outcomes.  The
`estimate` column returns the incidence rate ratio (`exp(beta)`) when the
family uses a log link.

```{r subgroup-glm, eval=FALSE}
subgroup_glm <- run_subgroup_analysis(
  data         = analysis_data,
  exposure     = "smoking_status",
  outcome      = "hospitalisation_count",
  subgroup_var = "sex",
  model_type   = "glm",
  family       = "poisson",
  covariates   = c("age", "bmi", "ethnic")
)
```

For overdispersed count outcomes, substitute `model_type = "negbin"` (no
`family` argument required).

```{r subgroup-negbin, eval=FALSE}
subgroup_nb <- run_subgroup_analysis(
  data         = analysis_data,
  exposure     = "smoking_status",
  outcome      = "hospitalisation_count",
  subgroup_var = "sex",
  model_type   = "negbin",
  covariates   = c("age", "bmi", "ethnic")
)
# estimate column = IRR = exp(beta); theta (dispersion) is reported per model
```

### Multiple subgroup variables

For multiple subgroup variables, use `run_multi_subgroup()` and visualize the
result with `plot_forest()`.

```{r subgroup-multi, eval=FALSE}
subgroup_all <- run_multi_subgroup(
  data         = analysis_data,
  exposure     = "smoking_status",
  subgroup_vars = c("sex", "age_group", "bmi_group"),
  model_type   = "cox",
  endpoint     = c("follow_up_time", "incident_status"),
  covariates   = c("age", "bmi", "ethnic")
)

plot_forest(
  subgroup_all,
  estimate_col = "estimate",
  lower_col    = "lower95",
  upper_col    = "upper95",
  null_value   = 1,
  log_scale    = TRUE
)
```

The same `model_type` and `family` arguments apply to `run_multi_subgroup()`.

### Interaction p-value notes

- For single-level binary subgroups the interaction p-value comes from the
  Wald test on the single interaction coefficient.
- For multi-level factors a likelihood-ratio test (LRT) is used: the full
  model with all interaction terms is compared against the main-effects model
  via `stats::anova()`.  Quasi-likelihood families fall back to an F-test.
- When the exposure has no variation within a subgroup level, the function
  returns `NA` for the effect estimate and issues a warning rather than
  stopping.

## Propensity score analysis {#propensity}

Propensity score methods can be used when comparing an exposure or treatment
group in observational data. The workflow is: estimate propensity scores,
perform matching or weighting, then fit the outcome model.

```{r ps-workflow, eval=FALSE}
ps_data <- estimate_propensity_score(
  data = analysis_data,
  treatment = "exposure_group",
  covariates = c("age", "sex", "bmi", "smoking_status", "ethnic"),
  method = "logistic"
)

matched_data <- match_propensity(
  data = ps_data,
  treatment = "exposure_group",
  ps_col = "ps",
  ratio = 1,
  caliper = 0.20
)
```

For IPTW analyses, use `calculate_weights()` followed by
`run_weighted_analysis()`.

```{r iptw-workflow, eval=FALSE}
weighted_data <- calculate_weights(
  data = ps_data,
  treatment = "exposure_group",
  ps_col = "ps",
  weight_type = "ATE",
  stabilized = TRUE
)

weighted_res <- run_weighted_analysis(
  data = weighted_data,
  exposure = "exposure_group",
  model_type = "cox",
  endpoint = c("follow_up_time", "incident_status"),
  weight_col = "weight"
)

weighted_res
```

## Mediation analysis {#mediation}

`run_mediation()` wraps regression-based mediation analysis and supports linear,
logistic, and Cox outcomes. It is useful when testing whether an exposure effect
is partly explained through an intermediate variable.

```{r mediation-basic, eval=FALSE}
med_res <- run_mediation(
  data = analysis_data,
  exposure = "smoking_status",
  mediator = "inflammatory_marker",
  outcome = "follow_up_time",
  outcome_type = "cox",
  endpoint = c("follow_up_time", "incident_status"),
  mediator_type = "continuous",
  covariates = c("age", "sex", "bmi", "ethnic"),
  exposure_levels = c(0, 1),
  boot = TRUE,
  boot_n = 1000
)

summary(med_res)
```

For screening several candidate mediators, use `run_multi_mediator()`.

```{r mediation-multi, eval=FALSE}
multi_med <- run_multi_mediator(
  data = analysis_data,
  exposure = "smoking_status",
  mediators = c("bmi", "crp", "hba1c"),
  outcome = "follow_up_time",
  outcome_type = "cox",
  endpoint = c("follow_up_time", "incident_status"),
  covariates = c("age", "sex", "ethnic")
)

multi_med
```

## Practical notes

- Keep the adjustment set consistent with the main analysis unless the advanced
  analysis has a specific causal estimand.
- Check subgroup sample sizes and event counts before interpreting interaction
  p-values.
- Inspect propensity score overlap and covariate balance before reporting
  matching or weighting results.
- For mediation analysis, define the temporal and biological ordering of
  exposure, mediator, and outcome before model fitting.
