# Bioinformatic and omics analyses {#bioinformatics}

This chapter introduces the omics-analysis layer in **UKBAnalytica**. The
current implementation focuses on plasma proteomics workflows, especially Olink
proteins from UK Biobank RAP analyses. Metabolomics support will be documented
in a future update.

The proteomics workflow is organized around three tasks:

1. association screening between proteins and clinical outcomes;
2. functional enrichment analysis of selected proteins;
3. STRING-based protein-protein interaction (PPI) network analysis.

In RAP projects, participant-level matrices should remain inside the approved
analysis environment. The examples below assume that users export only aggregate
regression results, enrichment summaries, and figures.

## Required packages

Proteomics enrichment and network analysis use optional Bioconductor and network
packages. Install them before running the examples.

```{r proteomics-install, eval=FALSE}
install.packages("BiocManager")
BiocManager::install(c("clusterProfiler", "AnnotationDbi", "org.Hs.eg.db"))

# Optional helper used by several PPI and enrichment wrappers
remotes::install_github("Hinna0818/TCMDATA")
```

## Proteomics input and identifier handling

RAP-exported Olink protein columns often appear as lower-case field-derived
names, for example `olink_instance_0.gdf15` or `olink_instance_0.il6`.
`protein_to_gene_symbol()` converts these identifiers to HGNC gene symbols for
enrichment and PPI analyses. It also supports UKB coding 143 style labels such
as `IL6;Interleukin-6` and multi-target labels such as `IL12A_IL12B`.

```{r proteomics-id-map, eval=FALSE}
library(UKBAnalytica)

protein_to_gene_symbol(c(
  "olink_instance_0.gdf15",
  "olink_instance_0.il6",
  "IL12A_IL12B;Interleukin-12"
))
```

If an upstream platform uses UniProt IDs or panel-specific identifiers, provide
a custom mapping table.

```{r proteomics-custom-map, eval=FALSE}
custom_map <- data.frame(
  protein = c("P05231", "P01375"),
  gene_symbol = c("IL6", "TNF")
)

protein_to_gene_symbol(
  proteins = c("P05231", "P01375"),
  from_type = "UNIPROT",
  mapping_table = custom_map
)
```

## Protein-outcome regression analysis

For incident disease outcomes, proteins can be screened with Cox proportional
hazards models using `runmulti_cox()`. Each protein is fitted as the main
exposure in a separate model, while the same adjustment covariates are applied
across all proteins.

```{r proteomics-cox, eval=FALSE}
protein_cols <- grep("^olink_instance_0\\.", names(train_data), value = TRUE)

cox_res <- runmulti_cox(
  data = train_data,
  main_var = protein_cols,
  endpoint = c("follow_up_time", "incident_status"),
  covariates = c(
    "age", "sex", "bmi", "smoking_status",
    "TDI", "ethnic", "education"
  )
)
```

`runmulti_cox()` returns one row per protein with `variable`, `coef`, `HR`,
`lower95`, `upper95`, `pvalue`, `n`, and `n_event`. Multiple-testing correction
should be applied before selecting candidate proteins.

```{r proteomics-cox-adjust, eval=FALSE}
cox_res$p_bh <- p.adjust(cox_res$pvalue, method = "BH")
cox_res$p_bonferroni <- p.adjust(cox_res$pvalue, method = "bonferroni")

cox_res <- merge(
  cox_res,
  protein_to_gene_symbol(cox_res, protein_col = "variable"),
  by.x = "variable",
  by.y = "protein",
  all.x = TRUE
)

sig_proteins <- subset(cox_res, p_bonferroni < 0.05)
```

For binary outcomes, the same structure can be used with `runmulti_logit()`,
followed by BH or Bonferroni correction of the returned p-values.

## Volcano visualization

`plot_regression_volcano()` accepts outputs from `runmulti_cox()` or
`runmulti_logit()`. Users specify the effect column, raw p-value column, and the
adjusted p-value column used for highlighting.

```{r proteomics-volcano, eval=FALSE}
volcano_plot <- plot_regression_volcano(
  cox_res,
  effect_col = "HR",
  p_col = "pvalue",
  adjusted_p_col = "p_bonferroni",
  label_col = "gene_symbol",
  significance_cutoff = 0.05,
  top_n_label_each = 5,
  null_effect = 1,
  x_lab = "Hazard ratio"
)

volcano_plot
```

For logistic regression results, set `effect_col = "OR"` and keep
`null_effect = 1`.

## Gene enrichment analysis

After selecting significant proteins, `run_protein_ora()` performs GO
over-representation analysis through `clusterProfiler::enrichGO()`. For
proteomics panels, the measured protein panel should usually be used as the
background universe instead of the whole genome.

```{r proteomics-go-bp, eval=FALSE}
sig_gene_symbols <- unique(na.omit(sig_proteins$gene_symbol))

go_bp <- run_protein_ora(
  proteins = sig_gene_symbols,
  universe = protein_cols,
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.20
)

go_table <- as.data.frame(go_bp$ora_result)
go_table[, c("ID", "Description", "GeneRatio", "p.adjust", "Count")]
```

The package provides light plotting helpers for enrichment results.

```{r proteomics-go-plot, eval=FALSE}
plot_enrichment_lollipop(
  go_bp,
  top_n = 10,
  plot_title = "GO biological process enrichment"
)
```

KEGG ORA can be run with `run_protein_kegg_ora()` when pathway-level annotation
is needed.

```{r proteomics-kegg, eval=FALSE}
kegg_res <- run_protein_kegg_ora(
  proteins = sig_gene_symbols,
  universe = protein_cols,
  pAdjustMethod = "BH"
)
```

## PPI network analysis

`get_protein_ppi()` maps proteins to gene symbols and retrieves STRING PPI
networks through `clusterProfiler::getPPI()`. The returned object stores both
the mapped symbols and the network, which helps users inspect identifier
conversion before downstream network analysis.

```{r proteomics-ppi, eval=FALSE}
ppi_res <- get_protein_ppi(
  proteins = sig_gene_symbols,
  taxID = 9606,
  network_type = "functional",
  output = "igraph"
)

ppi_res$gene_symbols
ppi_res$ppi
```

The network can then be filtered and annotated with topological metrics.

```{r proteomics-ppi-topology, eval=FALSE}
ppi_filtered <- subset_protein_ppi(
  ppi_res,
  score_cutoff = 0.70,
  rm_isolates = TRUE
)

ppi_metrics <- compute_protein_ppi_metrics(
  ppi_filtered,
  weight_attr = "score"
)

node_rank <- rank_protein_ppi_nodes(ppi_metrics)
head(node_rank$table)
```

## Fast greedy clustering of PPI modules

For module detection, `run_protein_ppi_fastgreedy()` applies igraph fast-greedy
community detection to the PPI graph. By default, it works on the largest
connected component and stores cluster labels in the `fast_greedy_cluster`
vertex attribute.

```{r proteomics-ppi-fastgreedy, eval=FALSE}
ppi_fg <- run_protein_ppi_fastgreedy(
  ppi_metrics,
  n_clusters = 4,
  largest_component = TRUE
)

fg_scores <- score_protein_ppi_clusters(
  ppi_fg,
  cluster_attr = "fast_greedy_cluster",
  min_size = 3
)

fg_scores
```

Cluster labels can be used for module-specific biological interpretation. A
common strategy is to run GO BP enrichment separately for proteins in each
cluster or to pass cluster-specific gene lists to `clusterProfiler::compareCluster()`.

```{r proteomics-ppi-cluster-enrichment, eval=FALSE}
library(igraph)

cluster_df <- data.frame(
  gene_symbol = V(ppi_fg)$name,
  cluster = V(ppi_fg)$fast_greedy_cluster
)

cluster_gene_list <- split(cluster_df$gene_symbol, cluster_df$cluster)

cluster_go <- clusterProfiler::compareCluster(
  geneCluster = cluster_gene_list,
  fun = "enrichGO",
  OrgDb = org.Hs.eg.db::org.Hs.eg.db,
  keyType = "SYMBOL",
  ont = "BP",
  pAdjustMethod = "BH"
)
```

When interpreting PPI clusters, keep the STRING score cutoff, disconnected
components, and cluster size threshold explicit in the methods section.

## Metabolomics analysis

Metabolomics workflows will be added in a future release. This section is
reserved for regression screening, multiple-testing correction, pathway
annotation, and metabolite-set enrichment examples.

## Practical recommendations

- Use the measured proteomics panel as the enrichment background when possible.
- Report both the raw p-value and the adjusted p-value method used for protein
  screening.
- Keep protein scaling and covariate adjustment consistent between training and
  validation analyses.
- For PPI analyses, record the STRING network type, score cutoff, filtering
  rules, and clustering algorithm.
- Export only aggregate summaries and publication figures from RAP projects.
