---
title: "Air pollution and incident arrhythmia: a UKBAnalytica demo case study"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  eval: true
  echo: true
  output: true
  warning: false
  message: false
  freeze: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This document demonstrates an end-to-end UKBAnalytica workflow using
`ukb_demo()` as a RAP-style raw table. The case study evaluates associations
between air pollution exposure and incident cardiac arrhythmia. The workflow
covers raw field extraction, phenotype definition, baseline preprocessing,
participant flow, baseline table generation, Cox regression, and sensitivity
analyses.

The demo data are synthetic and do not contain UK Biobank participant-level
source records. Results are for workflow illustration only.

## Load demo data

```{r load-packages}
if (requireNamespace("pkgload", quietly = TRUE) && dir.exists("../R")) {
  pkgload::load_all("..", quiet = TRUE)
}
library(UKBAnalytica)
library(data.table)

demo_raw <- ukb_demo()
dim(demo_raw)
names(demo_raw)[1:12]
```

## Specify variables from raw UKB field IDs

The analysis uses the following raw UKB fields:

- NO2: `p24016`, `p24018`, and `p24017`; `p24003` is used for the 2010-exposure
  sensitivity analysis.
- NOx: `p24004`.
- PM2.5: `p24006`.
- PM10: `p24019`; `p24005` is used for the 2010-exposure sensitivity analysis.
- Covariates: sex (`p31`), baseline date (`p53_i0`), age (`p21022`), BMI
  (`p21001_i0`), smoking (`p20116_i0`), drinking (`p20117_i0`), Townsend
  deprivation index (`p189`), ethnicity (`p21000_i0`), and education
  (`p6138_i0`).

## Define incident arrhythmia

The predefined `Arrhythmia` phenotype in UKBAnalytica captures broad cardiac
arrhythmia using ICD-10 codes `I44-I49` and procedure codes including `K57.6`,
`K59-K62`, `K64.1`, and `K72-K74` where available. Baseline history and
prospective incident events are handled separately.

```{r arrhythmia-definition}
disease_catalog <- get_predefined_diseases()
arrhythmia_def <- disease_catalog["Arrhythmia"]
arrhythmia_def

demo_surv <- build_survival_dataset(
  dt = demo_raw,
  disease_definitions = arrhythmia_def,
  prevalent_sources = c("ICD10", "OPCS4"),
  outcome_sources = c("ICD10"),
  baseline_col = "p53_i0",
  primary_disease = "Arrhythmia",
  censor_date = as.Date("2023-10-31"),
  show_flow = TRUE
)

table(demo_surv$outcome_status, useNA = "ifany")
participant_flow <- attr(demo_surv, "participant_flow")
participant_flow
```

`outcome_status = 1` indicates incident arrhythmia, `outcome_status = 0`
indicates censoring, and `NA` indicates baseline prevalent arrhythmia.

## Preprocess exposures and covariates

`preprocess_baseline()` maps raw UKB field names to standardized analysis
variables. Townsend deprivation index is added through `custom_mapping` because
it is not part of the default baseline variable set.

```{r preprocess}
exposure_vars <- c(
  "no2_2005", "no2_2006", "no2_2007", "no2_2010",
  "nox", "pm25", "pm10_2007", "pm10_2010"
)

covariates <- c("age", "sex", "tdi", "ethnicity", "education", "smoking", "drinking")

custom_mapping <- list(
  tdi = list(
    ukb_col = "p189",
    description = "Townsend deprivation index at recruitment"
  )
)

demo_prepared <- preprocess_baseline(
  df = demo_surv,
  variables = c(exposure_vars, covariates),
  custom_mapping = custom_mapping,
  missing_action = "keep"
)

demo_prepared[, c(exposure_vars, covariates), with = FALSE][1:6]
```

NO2 is summarized as the mean of 2005, 2006, and 2007 estimates for the main
analysis. The main PM10 exposure uses 2007 estimates. The 2010 NO2 and PM10
fields are retained for sensitivity analysis.

```{r derive-exposures}
demo_prepared[, no2_main := rowMeans(
  cbind(no2_2005, no2_2006, no2_2007),
  na.rm = TRUE
)]
demo_prepared[is.nan(no2_main), no2_main := NA_real_]

main_exposures <- c("no2_main", "nox", "pm25", "pm10_2007")
sensitivity_exposures_2010 <- c("no2_2010", "nox", "pm25", "pm10_2010")

main_exposures_z <- paste0(main_exposures, "_z")
sensitivity_exposures_2010_z <- paste0(sensitivity_exposures_2010, "_z")

for (var in unique(c(main_exposures, sensitivity_exposures_2010))) {
  demo_prepared[, (paste0(var, "_z")) := as.numeric(scale(get(var)))]
}

demo_prepared[, c(main_exposures, main_exposures_z), with = FALSE][1:6]
```

## Assemble the analytic cohort

The prospective cohort excludes participants with baseline prevalent arrhythmia,
missing outcome status, invalid follow-up time, or missing values in the main
exposures and adjustment covariates. The stepwise complete-case flow is retained
for reporting.

```{r cohort-assembly}
analysis_data <- demo_prepared[
  !is.na(outcome_status) &
    !is.na(outcome_surv_time) &
    outcome_surv_time > 0
]

complete_case_vars <- c(main_exposures_z, covariates)

analysis_data <- sensitivity_exclude_missing_covariates(
  data = analysis_data,
  covariates = complete_case_vars,
  stepwise = TRUE,
  verbose = TRUE
)

complete_case_flow <- attr(analysis_data, "complete_case_flow")
complete_case_flow

table(analysis_data$outcome_status)
```

## Baseline characteristics

Baseline characteristics are summarized by incident arrhythmia status. Air
pollutants are shown on their original scales, while Cox models use standardized
exposure variables.

```{r tableone}
table1 <- create_baseline_table(
  data = analysis_data,
  case_col = "outcome_status",
  factor_cols = c("sex", "ethnicity", "education", "smoking", "drinking"),
  continuous_cols = c("age", "tdi", main_exposures),
  test = TRUE
)

print(table1, quote = FALSE, noSpaces = TRUE)
```

## Cox regression

Each pollutant is evaluated in a separate Cox proportional hazards model,
adjusted for age, sex, Townsend deprivation index, ethnicity, education,
smoking, and drinking. Hazard ratios are interpreted per one-standard-deviation
increase in exposure.

```{r cox}
pollutant_labels <- data.frame(
  variable = main_exposures_z,
  pollutant = c("NO2", "NOx", "PM2.5", "PM10"),
  stringsAsFactors = FALSE
)

cox_results <- runmulti_cox(
  data = analysis_data,
  main_var = main_exposures_z,
  covariates = covariates,
  endpoint = c("outcome_surv_time", "outcome_status")
)

cox_results <- merge(cox_results, pollutant_labels, by = "variable", all.x = TRUE)
cox_results$p_bh <- p.adjust(cox_results$pvalue, method = "BH")
cox_results$p_bonferroni <- p.adjust(cox_results$pvalue, method = "bonferroni")
cox_results[order(cox_results$pvalue), ]
```

## Sensitivity analysis 1: exclude early events

To reduce potential reverse causation, early incident arrhythmia events are
excluded at 1-, 2-, and 4-year lag windows, and the adjusted Cox models are
refitted.

```{r sensitivity-lag}
fit_pollution_cox <- function(data, exposures, labels, lag_years = 0) {
  out <- runmulti_cox(
    data = data,
    main_var = exposures,
    covariates = covariates,
    endpoint = c("outcome_surv_time", "outcome_status")
  )
  out$lag_years <- lag_years
  merge(out, labels, by = "variable", all.x = TRUE)
}

lag_results <- lapply(c(0, 1, 2, 4), function(lag) {
  lag_data <- if (lag == 0) {
    analysis_data
  } else {
    sensitivity_exclude_early_events(
      data = analysis_data,
      endpoint = c("outcome_surv_time", "outcome_status"),
      n_years = lag,
      verbose = TRUE
    )
  }
  fit_pollution_cox(lag_data, main_exposures_z, pollutant_labels, lag_years = lag)
})

lag_results <- data.table::rbindlist(lag_results, use.names = TRUE)
lag_results[, p_bh := p.adjust(pvalue, method = "BH"), by = lag_years]
as.data.frame(lag_results[order(pollutant, lag_years)])
```

## Sensitivity analysis 2: 2010 exposure definition

NO2 and PM10 are redefined using 2010 exposure fields (`p24003` and `p24005`),
while NOx and PM2.5 are kept unchanged.

```{r sensitivity-2010-exposure}
pollutant_labels_2010 <- data.frame(
  variable = sensitivity_exposures_2010_z,
  pollutant = c("NO2_2010", "NOx", "PM2.5", "PM10_2010"),
  stringsAsFactors = FALSE
)

analysis_data_2010 <- sensitivity_exclude_missing_covariates(
  data = demo_prepared[
    !is.na(outcome_status) &
      !is.na(outcome_surv_time) &
      outcome_surv_time > 0
  ],
  covariates = c(sensitivity_exposures_2010_z, covariates),
  stepwise = FALSE,
  verbose = TRUE
)

cox_results_2010 <- fit_pollution_cox(
  data = analysis_data_2010,
  exposures = sensitivity_exposures_2010_z,
  labels = pollutant_labels_2010,
  lag_years = 0
)
cox_results_2010$p_bh <- p.adjust(cox_results_2010$pvalue, method = "BH")
cox_results_2010
```
