# Machine Learning Workflow {#machine-learning}

This chapter describes the recommended non-survival machine learning workflow
in **UKBAnalytica_v2**. The main user-facing function is `ukb_ml_workflow()`.
It is designed to keep the API simple while enforcing a standard analysis
structure:

1. Freeze the test set before model development.
2. Use only training and validation data for feature selection, tuning, and
   threshold learning.
3. Refit the final model on the training-development data.
4. Evaluate once on the frozen test set.

This chapter covers binary classification, multiclass classification,
continuous outcomes, and time-to-event prediction. For survival ML, use
`ukb_ml_survival_workflow()`; the legacy `ukb_ml_survival()` interface remains
available only for backward compatibility.

## What users usually need

For most projects, users only need one function:

```{r ml-minimal, eval=FALSE}
library(UKBAnalytica)

ml <- ukb_ml_workflow(
  outcome ~ age + sex + bmi + smoking + sbp,
  data = analysis_dt,
  model = "logistic",
  seed = 42
)

ml$final_test_metrics
head(ml$final_test_predictions)
```

This minimal call is useful for a quick baseline model. For a manuscript-ready
binary classification workflow, use an explicit train/validation/test split and
learn the classification threshold on validation data:

```{r ml-recommended-binary, eval=FALSE}
ml <- ukb_ml_workflow(
  outcome_status ~ age + sex + bmi + smoking + sbp + dbp,
  data = analysis_dt,
  model = "rf",
  split_params = list(
    split = "train_valid_test",
    train_ratio = 0.70,
    validation_ratio = 0.10,
    test_ratio = 0.20
  ),
  feature_select = "filter",
  feature_params = list(max_features = 50),
  tune = TRUE,
  tune_params = list(
    resampling = "validation",
    metric = "auc",
    param_grid = data.frame(
      num.trees = c(300, 500),
      mtry = c(3, 5),
      min.node.size = c(5, 10)
    )
  ),
  threshold_method = "youden",
  seed = 42
)

ml$final_test_metrics
head(ml$final_test_predictions)
```

The test set is not used during feature selection, tuning, or threshold
learning. It is used only by `ukb_ml_evaluate_test()` inside the final step.

## Prepare an ML-ready dataset

Start from an analysis-ready table. For incident disease prediction, this is
usually the wide table returned by `build_survival_dataset()`. Keep one row per
participant and include only variables available at prediction time.

```{r ml-prepare, eval=FALSE}
library(data.table)
library(UKBAnalytica)

ml_dt <- as.data.table(analysis_dt)[
  ,
  .(
    eid,
    outcome_status,
    age,
    sex,
    bmi,
    smoking,
    sbp,
    dbp,
    diabetes_history,
    hypertension_history
  )
]

ml_dt <- ukb_clean_missing(ml_dt)
ukb_snapshot(ml_dt, label = "ml_input")
```

Keep the outcome and predictors conceptually separated:

| Role | Example | Rule |
|:-----|:--------|:-----|
| Outcome | `outcome_status`, `diabetes`, `biomarker_value` | The target to predict |
| Predictors | `age`, `sex`, `bmi`, biomarkers | Must be known before the prediction time |
| ID | `eid` | Used to check overlap; do not use as a predictor |
| Leakage variables | follow-up time, diagnosis date after baseline | Do not include as predictors |

If you use `outcome_status` from `build_survival_dataset()`, make sure the
predictor set does not include follow-up information or post-baseline diagnosis
variables.

## Model choices

`ukb_ml_workflow()` supports the following non-survival models:

| `model` | Outcome type | Best use |
|:--------|:-------------|:---------|
| `"logistic"` | Binary | Baseline, interpretable classifier |
| `"linear"` | Continuous | Baseline regression |
| `"rf"` | Binary, multiclass, continuous | Robust nonlinear tabular model |
| `"xgboost"` | Binary, multiclass, continuous | Gradient boosting for stronger prediction |
| `"glmnet"` | Binary, multiclass, continuous | Regularized model and sparse features |
| `"svm"` | Binary, multiclass, continuous | Nonlinear boundaries in smaller datasets |
| `"nnet"` | Binary, multiclass, continuous | Simple neural network or multinomial model |
| `"rpart"` | Binary, multiclass, continuous | Lightweight decision tree baseline |
| `"naive_bayes"` | Binary, multiclass | Fast probabilistic classifier |

Outcome type is inferred automatically in common cases:

| Outcome values | Inferred `outcome_type` |
|:---------------|:------------------------|
| Two-level factor or character | `"binary"` |
| Numeric 0/1 | `"binary"` |
| Factor or character with 3+ levels | `"multiclass"` |
| Numeric with many values | `"continuous"` |

If the outcome is ambiguous, set `outcome_type` explicitly.

## Binary classification

Binary classification is the most common use case. Use a factor with the
negative class first and the positive class second when you want full control.

```{r ml-binary, eval=FALSE}
ml_dt$outcome_status <- factor(
  ml_dt$outcome_status,
  levels = c(0, 1),
  labels = c("control", "case")
)

ml <- ukb_ml_workflow(
  outcome_status ~ age + sex + bmi + smoking + sbp + dbp,
  data = ml_dt,
  model = "logistic",
  split_params = list(split = "train_valid_test"),
  tune = TRUE,
  tune_params = list(resampling = "validation", metric = "auc"),
  threshold_method = "youden",
  seed = 42
)

ml$final_test_metrics[c("auc", "accuracy", "sensitivity", "specificity", "f1")]
ml$threshold_result
```

For binary outcomes, `final_test_predictions` contains:

| Column | Meaning |
|:-------|:--------|
| `truth` | Observed class in the test set |
| `prob` | Predicted probability for the positive class |
| `pred_class` | Class after applying the fixed threshold |

## Continuous outcomes

Use `model = "linear"` for a baseline model or `model = "rf"`/`"xgboost"` for
nonlinear prediction.

```{r ml-continuous, eval=FALSE}
ml_reg <- ukb_ml_workflow(
  biomarker_value ~ age + sex + bmi + smoking + sbp,
  data = ml_dt,
  model = "linear",
  outcome_type = "continuous",
  split_params = list(split = "train_test", train_ratio = 0.75),
  threshold_method = "none",
  seed = 42
)

ml_reg$final_test_metrics[c("rmse", "mae", "rsquared")]
head(ml_reg$final_test_predictions)
```

Continuous predictions are reported in the `prediction` column.

## Multiclass outcomes

For multiclass classification, set `threshold_method = "none"` because a single
binary threshold is not meaningful.

```{r ml-multiclass, eval=FALSE}
ml_multi <- ukb_ml_workflow(
  disease_subtype ~ age + sex + bmi + smoking + sbp,
  data = ml_dt,
  model = "nnet",
  outcome_type = "multiclass",
  split_params = list(split = "train_valid_test"),
  tune = TRUE,
  tune_params = list(
    resampling = "validation",
    metric = "macro_f1",
    param_grid = data.frame(decay = c(0, 0.01, 0.05))
  ),
  threshold_method = "none",
  seed = 42
)

ml_multi$final_test_metrics[c("accuracy", "macro_f1", "logloss")]
head(ml_multi$final_test_predictions)
```

Multiclass predictions include `pred_class` plus one probability column per
class.

## When train/test files already exist

If another workflow has already created `train_df`, `valid_df`, and `test_df`,
register them with `ukb_ml_as_split()`. This function checks duplicated or
overlapping IDs when `id_col` is supplied.

```{r ml-manual-split, eval=FALSE}
split_obj <- ukb_ml_as_split(
  train_data = train_df,
  validation_data = valid_df,
  test_data = test_df,
  id_col = "eid",
  outcome = "outcome_status",
  outcome_type = "binary"
)

ml <- ukb_ml_workflow(
  outcome_status ~ age + sex + bmi + smoking + sbp,
  split = split_obj,
  model = "rf",
  threshold_method = "youden",
  seed = 42
)
```

This is useful when the cohort split must be fixed across multiple packages,
teams, or external validation analyses.

## Feature selection

Feature selection is optional and is always performed without touching the test
set.

| `feature_select` | Meaning | Notes |
|:-----------------|:--------|:------|
| `"none"` | Use all formula predictors | Recommended for small predictor sets |
| `"filter"` | Rank features by simple univariate signal | Fast baseline for many predictors |
| `"glmnet"` | Select nonzero regularized coefficients | Good for sparse linear signals |
| `"boruta"` | Boruta all-relevant feature selection | Requires `Boruta`; can be slower |

Example:

```{r ml-feature, eval=FALSE}
ml <- ukb_ml_workflow(
  outcome_status ~ .,
  data = ml_dt[, setdiff(names(ml_dt), "eid"), with = FALSE],
  model = "rf",
  feature_select = "glmnet",
  feature_params = list(max_features = 100),
  threshold_method = "youden",
  seed = 42
)

ml$feature_result$selected_features
```

Use `feature_select = "none"` when you already have a clinically defined
predictor set and do not want data-driven feature selection.

## Hyperparameter tuning

`ukb_ml_tune()` is called internally when `tune = TRUE`. The easiest standard
setup is validation tuning:

```{r ml-tune-validation, eval=FALSE}
ml <- ukb_ml_workflow(
  outcome_status ~ age + sex + bmi + smoking + sbp,
  data = ml_dt,
  model = "xgboost",
  split_params = list(split = "train_valid_test"),
  tune = TRUE,
  tune_params = list(
    resampling = "validation",
    metric = "auc",
    param_grid = data.frame(
      nrounds = c(50, 100),
      max_depth = c(2, 4),
      eta = c(0.05, 0.10)
    )
  ),
  threshold_method = "youden",
  seed = 42
)
```

If you do not create a validation set, use cross-validation inside the training
set:

```{r ml-tune-cv, eval=FALSE}
ml <- ukb_ml_workflow(
  outcome_status ~ age + sex + bmi + smoking + sbp,
  data = ml_dt,
  model = "rf",
  split_params = list(split = "train_test"),
  tune = TRUE,
  tune_params = list(resampling = "cv", folds = 5, metric = "auc"),
  threshold_method = "youden",
  seed = 42
)
```

Do not use test-set metrics to choose hyperparameters.

## Output object

`ukb_ml_workflow()` returns a list-like `ukb_ml_workflow` object:

| Element | Content |
|:--------|:--------|
| `split` | Frozen `ukb_ml_split` object |
| `feature_result` | Selected features and feature-selection metadata |
| `tune_result` | Candidate grid, best parameters, best development-set score |
| `threshold_result` | Binary threshold learned from validation, CV OOF, or train data |
| `final_model` | Refit final model |
| `final_test_metrics` | Test-set metrics computed once |
| `final_test_predictions` | Row-level test-set predictions |

Common inspection commands:

```{r ml-output, eval=FALSE}
print(ml)

ml$feature_result$selected_features
ml$tune_result$best_params
ml$threshold_result
ml$final_test_metrics

test_pred <- ml$final_test_predictions
```

Predict on a new RAP-side dataset with the final model:

```{r ml-predict-new, eval=FALSE}
new_prob <- predict(ml$final_model, newdata = new_participants, type = "prob")
new_class <- predict(ml$final_model, newdata = new_participants, type = "class")
```

For continuous outcomes, use `type = "response"`.

## SHAP interpretation

Use `ukb_shap()` directly on the workflow object. If `data` is not supplied,
the frozen test set stored in `ml$split$test` is used, so the interpretation is
aligned with the final reported performance.

```{r ml-shap, eval=FALSE}
shap <- ukb_shap(
  ml,
  sample_n = 500,
  nsim = 100,
  seed = 42
)

ukb_shap_summary(shap)
plot_shap_summary(shap)
```

You can also explain a final model on a specific external dataset:

```{r ml-shap-final, eval=FALSE}
external_shap <- ukb_shap(
  ml$final_model,
  data = external_validation_df,
  sample_n = 500,
  seed = 42
)
```

For multiclass models, choose the class probability to explain:

```{r ml-shap-multiclass, eval=FALSE}
shap_case <- ukb_shap(ml_multi, class_level = "case", sample_n = 500)
```

## Advanced: run the workflow step by step

Most users should use `ukb_ml_workflow()`. The lower-level functions are useful
when you need to save intermediate objects or compare multiple models on the
same frozen split.

```{r ml-stepwise, eval=FALSE}
split_obj <- ukb_ml_split_data(
  ml_dt,
  outcome = "outcome_status",
  split = "train_valid_test",
  seed = 42
)

features <- ukb_ml_feature_select(
  split_obj,
  outcome_status ~ age + sex + bmi + smoking + sbp,
  method = "filter",
  max_features = 50,
  seed = 42
)

tuned <- ukb_ml_tune(
  split_obj,
  formula = features$formula,
  model = "rf",
  resampling = "validation",
  metric = "auc",
  seed = 42
)

final <- ukb_ml_fit_final(
  split_obj,
  formula = features$formula,
  model = "rf",
  best_params = tuned$best_params,
  feature_spec = features,
  seed = 42
)

test_eval <- ukb_ml_evaluate_test(final, split_obj)
test_eval$metrics
```

## Legacy quick-model interface

`ukb_ml_model()`, `ukb_ml_cv()`, `ukb_ml_metrics()`, `ukb_ml_roc()`,
and `ukb_ml_compare()` remain available for backward compatibility, but they
now emit a deprecation warning. For new analyses intended for reporting, use
`ukb_ml_workflow()` because it makes the frozen-test design explicit.

`ukb_shap()` is not deprecated. It now works with both new workflow objects and
legacy `ukb_ml` objects.

```{r ml-legacy, eval=FALSE}
ml_rf <- ukb_ml_model(
  outcome_status ~ age + sex + bmi + smoking,
  data = ml_dt,
  model = "rf",
  task = "classification",
  split_ratio = 0.8,
  seed = 42
)

ukb_ml_metrics(ml_rf, ci = TRUE)
ukb_ml_roc(ml_rf)
```

## Survival machine learning

For time-to-event prediction, use `ukb_ml_survival_workflow()`. It follows the
same structure as `ukb_ml_workflow()`:

1. Freeze the survival test set.
2. Select features and tune hyperparameters using only training/validation data.
3. Refit the final model on training plus validation data.
4. Evaluate once on the frozen test set.

The simplest baseline uses a Cox model and requires only the `survival`
backend:

```{r ml-survival, eval=FALSE}
surv_ml <- ukb_ml_survival_workflow(
  Surv(outcome_surv_time, outcome_status) ~ age + sex + bmi + smoking,
  data = analysis_dt,
  model = "cox",
  split_params = list(
    split = "train_valid_test",
    train_ratio = 0.70,
    validation_ratio = 0.10,
    test_ratio = 0.20
  ),
  feature_select = "filter",
  feature_params = list(max_features = 50),
  tune = TRUE,
  tune_params = list(resampling = "validation"),
  evaluation_times = c(1, 3, 5, 10),
  seed = 42
)

surv_ml$final_test_metrics
head(surv_ml$final_test_predictions)
```

Supported survival models:

| `model` | Backend | Best use |
|:--------|:--------|:---------|
| `"cox"` | `survival` | Interpretable baseline survival model |
| `"coxnet"` | `glmnet` | Regularized high-dimensional Cox model |
| `"rsf"` | `randomForestSRC` | Nonlinear random survival forest |
| `"gbm_surv"` | `gbm` | Gradient-boosted Cox model |

The final metrics include Harrell's C-index and naive time-specific Brier
scores at `evaluation_times`. The Brier scores are not IPCW-adjusted; use them
as quick model checks, not as the only censoring-aware performance statistic.

Predict survival probabilities or risk scores from the final model:

```{r ml-survival-predict, eval=FALSE}
surv_prob <- predict(
  surv_ml,
  newdata = external_validation_df,
  times = c(1, 3, 5),
  type = "survival"
)

risk_score <- predict(surv_ml$final_model, newdata = external_validation_df, type = "risk")
```

If train/validation/test data already exist, register them explicitly:

```{r ml-survival-manual-split, eval=FALSE}
surv_split <- ukb_ml_survival_as_split(
  train_data = train_df,
  validation_data = valid_df,
  test_data = test_df,
  time = "outcome_surv_time",
  event = "outcome_status",
  id_col = "eid"
)

surv_ml <- ukb_ml_survival_workflow(
  Surv(outcome_surv_time, outcome_status) ~ age + sex + bmi + smoking,
  split = surv_split,
  model = "cox",
  evaluation_times = c(1, 3, 5),
  seed = 42
)
```

Lower-level survival workflow helpers mirror the non-survival API:
`ukb_ml_survival_split_data()`, `ukb_ml_survival_as_split()`,
`ukb_ml_survival_feature_select()`, `ukb_ml_survival_tune()`,
`ukb_ml_survival_fit_final()`, and `ukb_ml_survival_evaluate_test()`.

`ukb_ml_survival()` now emits a deprecation warning. Use
`ukb_ml_survival_workflow()` for new analyses.

## Practical checklist

Before reporting ML results, check the following:

1. The test set was created before feature selection and tuning.
2. `eid`, diagnosis dates, follow-up time, and post-baseline variables are not
   used as predictors.
3. The outcome type is correct (`binary`, `multiclass`, or `continuous`).
4. Binary positive class is the second factor level if you set factor levels
   manually.
5. Hyperparameters and thresholds are chosen on train/validation or CV only.
6. `final_test_metrics` are reported as final frozen-test performance.
7. For survival ML, `outcome_surv_time` and `outcome_status` are used only as
   the left-hand side of `Surv(time, event)`, not as predictors.

## Optional packages

Install only the model backends you plan to use. UKBAnalytica_v2 checks optional
packages only when the selected model or helper needs them. It does not install
extra ML dependencies unless you explicitly opt in:

```{r ml-auto-install, eval=FALSE}
options(UKBAnalytica.auto_install_ml = TRUE)
```

The default is `FALSE`, which is safer on RAP and HPC systems because users can
control the package library and CRAN/Bioconductor mirror themselves.

```{r ml-install, eval=FALSE}
install.packages(c(
  "ranger",      # random forest
  "xgboost",     # gradient boosting
  "glmnet",      # regularized regression and glmnet feature selection
  "e1071",       # SVM and naive Bayes
  "nnet",        # neural network / multinomial regression
  "rpart",       # decision tree
  "Boruta",      # Boruta feature selection
  "fastshap",    # SHAP explanations
  "pROC",        # legacy ROC helpers
  "survival",    # Cox and survival metrics
  "gbm",         # boosted survival model
  "randomForestSRC"  # random survival forest
))
```

`logistic` and `linear` models use base R and do not require extra model
backend packages. `naive_bayes` reuses `e1071`, so no separate package is
needed if you already installed it for SVM. For survival ML, `model = "cox"`
uses `survival`, `model = "coxnet"` uses `glmnet`, `model = "gbm_surv"` uses
`gbm`, and `model = "rsf"` uses `randomForestSRC`.
