# Proteomics Enrichment Analysis {#proteomics-enrichment}

This chapter introduces the proteomics enrichment module in **UKBAnalytica**.
It is designed for UK Biobank RAP Olink proteomics data, where protein columns
often appear as lower-case names with an Olink prefix such as
`olink_instance_0.eno2` or `olink_instance_0.il6`.

The workflow has three steps:

1. Convert RAP-exported protein identifiers to gene symbols.
2. Run GO or KEGG over-representation analysis (ORA).
3. Visualize enrichment results with the TCMDATA plotting helpers.

## Required packages

The enrichment functions rely on optional Bioconductor packages. Install them
before running this chapter.

```{r proteomics-install, eval=FALSE}
# CRAN helper
install.packages("BiocManager")

# Bioconductor packages used by UKBAnalytica enrichment helpers
BiocManager::install(c("clusterProfiler", "AnnotationDbi", "org.Hs.eg.db"))

# Optional plotting package used by plot_go_ora_bar() and plot_enrichment_lollipop()
devtools::install_github("Hinna0818/TCMDATA")
```

## Supported protein name formats

`protein_to_gene_symbol()` is designed to handle the most common Olink naming
styles encountered in UKB RAP workflows:

1. RAP-exported column names such as `olink_instance_0.eno2`.
2. UKB coding 143 style strings such as `IL6;Interleukin-6`.
3. Multi-target Olink labels such as `IL12A_IL12B`, which will be expanded to
   separate gene symbols.

## Convert RAP-exported Olink names to gene symbols

Use `protein_to_gene_symbol()` when you already have a vector of selected
proteins, for example after filtering differential proteins or selecting a
protein panel of interest.

```{r proteomics-map-basic, eval=FALSE}
library(UKBAnalytica)

proteins <- c(
  "olink_instance_0.eno2",
  "olink_instance_0.eno3",
  "olink_instance_0.enoph1",
  "olink_instance_0.enox2"
)

protein_to_gene_symbol(proteins)
```

Expected output:

| protein | gene_symbol | mapping_source |
|:--------|:------------|:---------------|
| ENO2 | ENO2 | input_symbol |
| ENO3 | ENO3 | input_symbol |
| ENOPH1 | ENOPH1 | input_symbol |
| ENOX2 | ENOX2 | input_symbol |

You can also pass a data.frame together with `protein_col`.

```{r proteomics-map-dataframe, eval=FALSE}
res_df <- data.frame(
  protein = c(
    "olink_instance_0.il6",
    "olink_instance_0.tnf",
    "IL12A_IL12B;Interleukin-12"
  ),
  logFC = c(0.82, 0.76, 0.55)
)

protein_to_gene_symbol(res_df, protein_col = "protein")
```

In this example, `IL12A_IL12B` will be expanded into two rows: `IL12A` and
`IL12B`.

## GO over-representation analysis

`run_protein_ora()` converts protein identifiers to gene symbols and runs GO ORA
through `clusterProfiler::enrichGO()`.

```{r proteomics-go, eval=FALSE}
library(UKBAnalytica)

inflam_proteins <- c(
  "olink_instance_0.il6",
  "olink_instance_0.tnf",
  "olink_instance_0.il1b",
  "olink_instance_0.ifng",
  "olink_instance_0.ccl2",
  "olink_instance_0.ccl3",
  "olink_instance_0.ccl4",
  "olink_instance_0.cxcl10",
  "olink_instance_0.il10",
  "olink_instance_0.il6r",
  "olink_instance_0.tlr4",
  "olink_instance_0.icam1"
)

go_res <- run_protein_ora(
  proteins = inflam_proteins,
  ont = "ALL"
)

go_res$gene_symbols
as.data.frame(go_res$ora_result)[, c("ID", "Description", "p.adjust", "Count")]
```

If you want to draw the ontology-colored GO bar plot, use `ont = "ALL"` so the
result contains the `ONTOLOGY` column required by `plot_go_ora_bar()`.

## KEGG over-representation analysis

`run_protein_kegg_ora()` automatically converts gene symbols to Entrez IDs and
then calls `clusterProfiler::enrichKEGG()`.

```{r proteomics-kegg, eval=FALSE}
kegg_res <- run_protein_kegg_ora(
  proteins = inflam_proteins
)

kegg_res$entrez_ids
as.data.frame(kegg_res$ora_result)[, c("ID", "Description", "p.adjust", "Count")]
```

By default the function uses the human KEGG code `hsa`. If you are working with
very small protein sets and obtain no results, you can temporarily reduce
`minGSSize` for exploratory checks.

```{r proteomics-kegg-small, eval=FALSE}
kegg_res_small <- run_protein_kegg_ora(
  proteins = inflam_proteins,
  minGSSize = 2
)
```

## Use a custom background universe

When you perform enrichment on a subset of proteins from a measured Olink panel,
it is often better to use the assayed proteins as the background rather than the
whole genome.

```{r proteomics-universe, eval=FALSE}
# Example: all assayed Olink proteins in your matrix
all_proteins <- grep("^olink_instance_0\\.", names(protein_df), value = TRUE)

go_res <- run_protein_ora(
  proteins = inflam_proteins,
  universe = all_proteins,
  ont = "ALL"
)

kegg_res <- run_protein_kegg_ora(
  proteins = inflam_proteins,
  universe = all_proteins
)
```

## Plot enrichment results

The package provides two thin wrappers around plotting functions from
**TCMDATA**.

```{r proteomics-plot, eval=FALSE}
# GO bar plot
plot_go_ora_bar(
  go_res,
  top_n = 5,
  plot_title = "GO enrichment of inflammatory proteins"
)

# General lollipop plot for enrichment results
plot_enrichment_lollipop(
  go_res,
  top_n = 10,
  plot_title = "GO ORA lollipop plot"
)
```

`plot_enrichment_lollipop()` also works on KEGG enrichment results.

```{r proteomics-kegg-plot, eval=FALSE}
plot_enrichment_lollipop(
  kegg_res,
  top_n = 10,
  plot_title = "KEGG ORA lollipop plot"
)
```

## Use a custom mapping table

If your upstream proteomics workflow outputs UniProt IDs, panel-specific names,
or another naming convention, you can provide a custom mapping table first and
then let `clusterProfiler::bitr()` handle the remaining unmatched identifiers.

```{r proteomics-custom-map, eval=FALSE}
custom_map <- data.frame(
  protein = c("P05231", "P01375"),
  gene_symbol = c("IL6", "TNF")
)

protein_to_gene_symbol(
  proteins = c("P05231", "P01375"),
  from_type = "UNIPROT",
  mapping_table = custom_map
)
```

## Run STRING PPI analysis for proteomics hits

The package also provides a thin PPI workflow built on top of
`clusterProfiler::getPPI()` and the downstream network-analysis functions in
**TCMDATA**.

Start by retrieving a STRING network from your protein list.

```{r proteomics-ppi-retrieve, eval=FALSE}
ppi_res <- get_protein_ppi(
  proteins = inflam_proteins,
  taxID = 9606
)

ppi_res$gene_symbols
ppi_res$ppi
```

`get_protein_ppi()` returns both the mapped gene symbols and the retrieved PPI
network, so users can check the identifier conversion before moving to network
analysis.

## Filter the PPI network and compute topology

Use `subset_protein_ppi()` to remove low-confidence edges and optionally keep
only the most connected nodes. Then compute node-level topological metrics with
`compute_protein_ppi_metrics()`.

```{r proteomics-ppi-metrics, eval=FALSE}
ppi_filtered <- subset_protein_ppi(
  ppi_res,
  score_cutoff = 0.7,
  n = 100
)

ppi_metrics <- compute_protein_ppi_metrics(
  ppi_filtered,
  weight_attr = "score"
)
```

To prioritize hub proteins, run integrated centrality ranking.

```{r proteomics-ppi-rank, eval=FALSE}
rank_res <- rank_protein_ppi_nodes(ppi_metrics)

head(
  rank_res$table[, c(
    "name", "degree", "betweenness_w", "closeness_w",
    "MCC", "MNC", "EPC", "Score_network", "Rank_network"
  )]
)
```

## Detect PPI modules

Three clustering wrappers are included: Louvain, MCODE, and MCL.

```{r proteomics-ppi-cluster, eval=FALSE}
# Louvain
ppi_louvain <- run_protein_ppi_louvain(ppi_metrics)
score_protein_ppi_clusters(ppi_louvain)

# MCODE
ppi_mcode <- run_protein_ppi_mcode(ppi_metrics)
get_protein_mcode_res(ppi_mcode, only_clusters = TRUE)

# MCL
ppi_mcl <- run_protein_ppi_mcl(ppi_metrics)
score_protein_ppi_clusters(ppi_mcl, cluster_attr = "MCL_cluster")
```

## Assess network robustness

`run_protein_ppi_robustness()` maps the target proteins to node names in the
PPI graph and then calls `TCMDATA::ppi_knock()` to evaluate network disruption.

```{r proteomics-ppi-robustness, eval=FALSE}
robustness_res <- run_protein_ppi_robustness(
  ppi_metrics,
  targets = c("olink_instance_0.il6", "olink_instance_0.tnf")
)

robustness_res$targets
robustness_res$robustness$Summary
robustness_res$robustness$Total_Score
robustness_res$robustness$Total_Pvalue
```

## Practical notes

1. These enrichment helpers are currently designed for human proteomics data.
2. RAP-exported Olink names are normalized automatically, so users do not need
   to manually strip `olink_instance_0.` prefixes.
3. GO uses gene symbols directly, while KEGG internally converts them to Entrez
   IDs.
4. For ORA, the choice of background universe has a strong effect on the final
   results and should match the assayed protein space whenever possible.
5. The PPI workflow requires both **clusterProfiler** and **TCMDATA** to be
   installed.
6. Downstream PPI helpers expect an `igraph` network, so keep `output =
   "igraph"` in `get_protein_ppi()` if you plan to continue with topology,
   clustering, or robustness analysis.